---
title: "Using recoup with non-continuous regions (RNA-Seq)"
author: "Panagiotis Moulos"
date: "`r BiocStyle::doc_date()`"
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Using recoup with non-continuous regions (RNA-Seq)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

Construction of genomic coverage profiles in non-continuous regions
===================================================================

This tutotial demonstrates a few examples of how you can use `recoup` to create
average genomic cpverage profiles and heatmaps from short read sequences for 
genomic regions who are __non-continuous__. Such regions are for example genes 
of an organism where one would like to create the profile of RNA-Seq spliced 
reads. If the ChIP-Seq mode was to be applied in this case, the resulting 
profiles, although averaged, would be sparse and would simply reflect a Genome 
Browser image. This is also useful but can be achieved much better with other 
packages like [ggbio](http://www.tengfei.name/ggbio/). Most parameters are 
similar as in the ChIP-Seq case. The main parameter that changes is the `type` 
and then there are a few others which control some aspects of coverage 
generation. For further details, have a look at the man page of the `recoup` 
function.

The RNA-Seq plots, apart from expression averaging, are also very useful for
quality control purposes, e.g. RNA-degradation checks. This is not something
novel and explained very well in other packages like [ngs.plot](https://github.com/shenlab-sinai/ngsplot). What is novel here, is the
use of modern R graphics mechanisms in order to easily create and display data
categories.

## Get some data

The small dataset attached to the package does not contain RNA-Seq data. For the
rest of this tutorial, it is supposed that you have uncompressed the tar archive
presented in the introductory page in a directory called 
`/home/me/recoup_tutorial`. We will use the gene expression data located in `/home/me/recoup_tutorial/rnaseq`. Have a look at 
the [introductory tutorial page](A_recoup_intro.html) for a short description of 
the test data.

## Running recoup with RNA-Seq data

As mentioned in the [ChIP-Seq profiles tutorial](recoup_chipseq.html), the 
`recoup` function can be run either using a text __tab__ delimited file or a 
list of lists, which we are using in this case:

```{r echo=TRUE, eval=TRUE}
library(recoup)

test.path <- "/home/me/recoup_tutorial/rnaseq"

rna.input <- list(
    list(
        id="WT",
        name="WT",
        file=file.path(test.path,"WT.bam"),
        format="bam"
    ),
    list(
        id="Set8KO",
        name="Set8KO",
        file=file.path(test.path,"Set8KO.bam"),
        format="bam"
    )
)
```

All the following examples will use all your cores if you are using a multicore
system in Linux (not in Windows as the `parallel` package is not supported). If
you want to restrict the number of cores, use the `rc` argument. In real world
examples, you _should_ most of the times restrict the number of cores used as
the memory footprints are quite high in R parallel computations, unless of 
course you have a lot of memory to spare. We are working to improve that at 
least for `recoup`!

### RNA-Seq (gene body) profiles

When `type="rnaseq"`, the region is `genebody` by default, so there is no reason
to specify it. Plotting is done in the same way as other packages: introns are
removed and exons are placed side-by-side to create a continuous profile. The
region binning parameters apply here too, with the most important being the
action to be taken with reads that span splicing junctions. Have a look at the 
man page for further details.

This tutorial does not use custom genomic regions. It uses the local annotation
for mouse genes, version mm9. If you wish to run the examples, make sure that
you have execute `buildAnnotationStore` with ensembl mouse genes at least. This
is not required as annotation can be created on the fly but takes much more 
time.

The following will produce an average profile of RNA-Seq reads over gene bodies
with no design applied and linear normalization across samples (Hint: `log2` 
scale produced much more smooth profiles than natural scale in the RNA-Seq 
case).

```{r echo=TRUE, eval=TRUE}
test <- recoup(
    input=rna.input,
    type="rnaseq",
    genome="mm9",
    flank=c(1000,1000),
    binParams=list(flankBinSize=50,regionBinSize=100),
    selector=NULL,
    preprocessParams=list(normalize="linear"),
    plotParams=list(plot=FALSE,signalScale="log2")
)

plot(test$plots$profile)
draw(test$plots$heatmap,gap=grid::unit(1,"cm"))
```

Like the previous one, but we are making use of the `selector` argument which 
allows some easy subsetting of predefined genomic regions. We also apply 
downsampling normalization and ordering.

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
test <- recoup(
    input=rna.input,
    design=NULL,
    type="rnaseq",
    genome="mm9",
    flank=c(1000,1000),
    orderBy=list(what="sum1",order="descending"),
    binParams=list(flankBinSize=50,regionBinSize=100),
    selector=list(biotype=c("protein_coding","lincRNA","pseuodgene")),
    preprocessParams=list(normalize="downsample"),
    plotParams=list(plot=FALSE,signalScale="log2")
)

plot(test$plots$profile)
draw(test$plots$heatmap,gap=grid::unit(1,"cm"))
```

No design, but k-means clustering:

```{r echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.height=8}
test <- recoup(
    input=rna.input,
    design=NULL,
    type="rnaseq",
    genome="mm9",
    flank=c(1000,1000),
    binParams=list(flankBinSize=50,regionBinSize=200),
    selector=NULL,
    kmParams=list(k=4,reference="WT"),
    plotParams=list(plot=FALSE,signalScale="log2",heatmap=FALSE)
)

plot(test$plots$profile)
```

__Note__: In all cases, the ordering of the profiles as well as well as the 
elements (genomic areas) belonging to a group wheh using design files for 
grouping, can be acquired either from the `rownames` of the `profile` member
of the `recoup$data` output list object, or from the `design` member of the 
`recoup` output object. From the `design` member you can also extract the number
of elements in each group, which is very useful especially in the case of 
applying k-means clustering to the profiles.

## R session information

```{r echo=FALSE}
sessionInfo()
```

